Meta-analysis of composite outcomes in clinical trials

### 2025-01-29 MSc_poster_template.Rmd
Feed back for background addressed. Coding disparities for loading data addressed

# Background and Problem
Composites are commonly used in clinical trials as the primary outcome. When misused, they can reduce the power to detect treatment effect. This was seen in EMERGE, [^1] a randomised clinical trial investigating the effects of early initiation of Metformin treatment on developing gestational diabetes. The composite was created using three measures of glycemic control; blood glucose levels of 5.1mmol/L or greater at gestation weeks 32 or 38, or initiation of Insulin treatment.

@@

This composite involves dichotomising blood glucose levels, greatly reducing power. Therefore, using data from EMERGE, the objectives of this project are:

1.	To perform a meta-analysis of the primary outcome using permutations of the composite test statistic, without dichotomising glucose variables

2.	To compare various ways of creating the composite test statistic, and their power

3.	To compare the permutation method with other approaches to analyzing composite outcomes

@@

# Data Sources and Datasets

The syntax in this poster template and the `posterdown` package uses the same workflow approach as the R Markdown you know and love.

@@

# Descriptive Statistics of Datasets

Usually you want to have a nice table displaying some important results that you have calculated. In `posterdown` this is as easy as using the `kable` table formatting you are probably use to as per typical R Markdown formatting.

You can reference tables like so: Table \@ref(tab:mytable). Some basic summaries of the dataset are below:

```{r mytable, out.width='80%'}
knitr::kable(iris[1:10, 1:4], caption = 'Table caption.',align = 'c',"html") %>%
  kable_styling(font_size = 20)
```

Figure \@ref(fig:standard-plot), and Figure \@ref(fig:morefigs) below show the patterns in our dataset. Make sure that all the details in your plots will be legible when printed (legend text, axis text, and any labels)

```{r standard-plot, out.width='80%', fig.align='center', fig.cap='Great figure!', fig.height=4}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()
```

>>>>

# Background and Problem
Our project centres on  data from EMERGE, [^1] which was a randomised clinical trial investigating the effects of early initiation of Metformin treatment on developing gestational diabetes. The primary outcome of EMERGE was a composite created using three variables; blood glucose levels of 5.1mmol/L or greater at gestation weeks 32 or 38, or initiation of Insulin treatment.

@@

```{r}

composite_metformin <- round(sum(tmpDat$compositeYN[tmpDat$Group == "Metformin"] == "Yes", na.rm = TRUE) /
                               sum(!is.na(tmpDat$compositeYN[tmpDat$Group == "Metformin"])) * 100, 2)
composite_placebo <- round(sum(tmpDat$compositeYN[tmpDat$Group == "Placebo"] == "Yes", na.rm = TRUE) /
                             sum(!is.na(tmpDat$compositeYN[tmpDat$Group == "Placebo"])) * 100, 2)
EMERGE_composite_p_value<-round(prop.test(x = c(sum(tmpDat$compositeYN[tmpDat$Group == "Metformin"] == "Yes", na.rm = T),
                                                sum(tmpDat$compositeYN[tmpDat$Group== "Placebo"]== "Yes", na.rm = T)),
                                          n = c(sum(!is.na(tmpDat$compositeYN[tmpDat$Group == "Metformin"])), sum(!is.na(tmpDat$compositeYN[tmpDat$Group == "Placebo"])))
)$p.value, 2)
EMERGE_background_stats<-c("Metformin" = composite_metformin, "Placebo" = composite_placebo,
                           "P-value" = EMERGE_composite_p_value)

EMERGE_background_stats <- data.frame(t(as.matrix(EMERGE_background_stats)))
knitr::kable(EMERGE_background_stats, caption = 'EMERGE *p*-values', align = 'c', format = "html") %>%
  kable_styling(font_size = 20) %>%
  add_header_above(c("Primary Composite Outcome N/total (%)" = 2, " " = 1))
```

While the results of EMERGE were insignificant, the composite endpoint involved dichotomising blood glucose levels, greatly reducing power. Therefore, using data from EMERGE, the objectives of this project are:

  1.	to investigate the potential of meta-analysis to improve power over standard analysis of composite endpoints when some of the components of the composite are continuous and some are binary

2.	To compare various ways of creating the composite test statistic which is used in the meta-analysis, and their power

3.	To compare the permutation-based meta analysis approach we have proposed here to other alternatives to standard analysis of composite endpoint

@@

@@

```{r standard-plot, out.width='80%', fig.align='center', fig.cap='Great figure!', fig.height=4}
# Load the necessary libraries
library(ggplot2)
library(hrbrthemes)
library(readr)
library(dplyr)
library(reshape2)

# Create a binary column for Insulin
tmpDat$binary_insulin <- 0
tmpDat$binary_insulin[tmpDat$InsulinYN == "Yes"] <- 1

# Proportion test for Insulin usage between groups
StudyInsulinTest <- prop.test(
  x = c(
    sum(tmpDat$InsulinYN[tmpDat$Group == "Metformin"] == "Yes", na.rm = TRUE),
    sum(tmpDat$InsulinYN[tmpDat$Group == "Placebo"] == "Yes", na.rm = TRUE)
  ),
  n = c(
    sum(tmpDat$Group == "Metformin"),
    sum(tmpDat$Group == "Placebo")
  )
)

# Perform t-tests for binary_insulin, gw32, and gw38
insulin_t <- t.test(tmpDat$binary_insulin ~ tmpDat$Group)
gw32_t <- t.test(tmpDat$gw32 ~ tmpDat$Group)
gw38_t <- t.test(tmpDat$gw38 ~ tmpDat$Group)

# Calculate composite scores
y_squared_composite <- (insulin_t$statistic^2) + (gw32_t$statistic^2) + (gw38_t$statistic^2)
study_non_squared_composite <- insulin_t$statistic + gw32_t$statistic + gw38_t$statistic

# Visualization: Proportion of Insulin Usage by Group
ggplot(tmpDat, aes(x = Group, fill = InsulinYN)) +
  geom_bar(position = "fill") +
  labs(
    title = "Proportion of Insulin Usage by Group",
    y = "Proportion",
    x = "Group"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()

# Visualization: Boxplots for gw32 and gw38 with cutoff line at 5.1 and p-values
ggplot(tmpDat, aes(x = Group, y = gw32, fill = Group)) +
  geom_boxplot() +
  geom_hline(yintercept = 5.1, linetype = "dashed", color = "red") +
  annotate("text", x = 1.5, y = max(tmpDat$gw32, na.rm = TRUE), 
           label = paste("p-value:", round(gw32_t$p.value, 4)), 
           color = "blue", size = 4) +
  labs(
    title = "Gestational Week 32 by Group",
    y = "Week 32",
    x = "Group",
    caption = "Dashed red line represents the cutoff at 5.1"
  ) +
  theme_minimal()

ggplot(tmpDat, aes(x = Group, y = gw38, fill = Group)) +
  geom_boxplot() +
  geom_hline(yintercept = 5.1, linetype = "dashed", color = "red") +
  annotate("text", x = 1.5, y = max(tmpDat$gw38, na.rm = TRUE), 
           label = paste("p-value:", round(gw38_t$p.value, 4)), 
           color = "blue", size = 4) +
  labs(
    title = "Gestational Week 38 by Group",
    y = "Week 38",
    x = "Group",
    caption = "Dashed red line represents the cutoff at 5.1"
  ) +
  theme_minimal()

# Calculate and display correlation matrix
correlation_vars <- tmpDat %>%
  select(gw32, gw38, binary_insulin) 
correlation_matrix <- cor(correlation_vars, use = "complete.obs", method = "pearson")

# Plot correlation matrix as a heatmap
correlation_melted <- melt(correlation_matrix)

ggplot(correlation_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(
    title = "Correlation Matrix (Pearson)",
    x = "Variables",
    y = "Variables"
  ) +
  theme_minimal()
